---
title: "Data 607 Week 7 Assignment"
author: "by Dmitriy Vecheruk"
date: "15 October 2016"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    theme: united
    code_folding: show
    
---

## Introduction
  
This week's assignment is about reading three file formats into R:
  
- HTML
- XML
- JSON
  
I have manually created the three files describing the authors and titles of the books on decision making and have uploaded them on GitHub.


## Reading HTML Table
  
In a document with a simple structure, reading an HTML table is easy using the `readHTMLTable` function.
```{r}
library(XML)
library(knitr)
library(RCurl)

url_HTML = getURL("https://raw.githubusercontent.com/datafeelings/data607/master/week_7/books.html", ssl.verifypeer = FALSE)

books_HTML = readHTMLTable(url_HTML, 
                           header = T, 
                           which=1, 
                           isHTML = TRUE, 
                           stringsAsFactors = F)
```
The output is directly converted to a dataframe
```{r}
kable(books_HTML)
```


## Reading XML 
  
First, parse the XML tree and have a look at the document structure
```{r}

url_XML = getURL("https://raw.githubusercontent.com/datafeelings/data607/master/week_7/books.xml", ssl.verifypeer = FALSE)

books_XML = xmlParse(url_XML)

book_child = xmlChildren(books_XML)

head(book_child)

```
We can see that each element that we need has a list of tags with id, title, authors, year, and ISBN.  
Now we can populate a dataframe with the contents of each tag.
```{r}
books_XML_df = data.frame()

books_XML_df = cbind(xpathSApply(books_XML, "//id", xmlValue),
                     xpathSApply(books_XML, "//title", xmlValue),
                     xpathSApply(books_XML, "//authors", xmlValue),
                     xpathSApply(books_XML, "//year", xmlValue),
                     xpathSApply(books_XML, "//ISBN", xmlValue))

books_XML_df = data.frame(books_XML_df)
names(books_XML_df) = c("id", "title", "authors", "year", "ISBN")

kable(books_XML_df)
```


## Reading a JSON 
  
Similarly to the XML file, the JSON first should be parsed first, which is easily accomplished by the `jsonlite package`.
However, as I have picked a different node structure than in the previous formats, the parser interprets the node of "authors" as a list of character vectors (as one of the records is an array of two authors).

```{r}
library(jsonlite)

url_JSON = getURL("https://raw.githubusercontent.com/datafeelings/data607/master/week_7/books.json", ssl.verifypeer = FALSE)

books_JSON = fromJSON(txt = url_JSON)

str(books_JSON)
```

Forcing this file into a dataframe results in an incorrect data frame structure
```{r}
books_JSON_df = as.data.frame(books_JSON)

kable(books_JSON_df)
```
  
One solution is to collapse any embedded lists into characters first so that the "authors" column could be turned into a simple character vector.

```{r}
# Function to collapse any list longer than 1 into a character vector
collapselist = function(x) { 
  x = ifelse(length(unlist(x))>1,
             yes = paste(unlist(x),collapse = ", "),
             no = as.character(x))
}

books_JSON$Autors = sapply(books_JSON$Autors,FUN = collapselist)

books_JSON_df = as.data.frame(books_JSON)

kable(books_JSON_df)

```
  
Now the data frame is correct.
