---
title: "DATA 607: Week 3 Assignment"
author: "by Dmitriy Vecheruk"
date: "12 September 2016"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    theme: united
    code_folding: show
---

  

# String processing 

This assignment is focused on using the stringr package and base R tools to extract data from messy text strings:  
- Problem 3 is about extracting names  
- Problem 4 is about more complex regular expressions  
- Problem 9 is a riddle  

## Problem 3

This example introduces the `stringr` package and extracts the names of characters from the raw data.  
The assignment consists of three parts with the final goal to provide a string of character names conforming to the order `first_name last_name` 

### The introductory example  



```{r}
# Code from the introductory example
# Source: Munzert, S. et. al. (2015). Automated Data Collection with R - A Practical Guide to Web Scraping and Text Mining. Wiley.

  raw.data <- "555-1239Moe Szyslak(636) 555-0113Burns, C. Montgomery555-6542Rev. Timothy Lovejoy555 8904Ned Flanders636-555-3226Simpson, Homer5553642Dr. Julius Hibbert"
  
  
  library(stringr)
  name <- unlist(str_extract_all(raw.data, "[[:alpha:]., ]{2,}"))

```

The resulting string looks like this  
```{r}
name
```


### 3(b) Identify records containing titles

```{r}
names_w_titles = str_detect(name, "Rev.|Dr.")
```

Here are the names again and the correspondent logical values for titles

```{r}
  name
  names_w_titles

```

### 3(a) Rearrange to correct order  

The goal is to rearrange the following string so that all names follow the order  `first_name last_name`.

```{r echo=FALSE}
  name

```


First we remove the titles from the names using the output of 3(b)

```{r}
name1 = str_replace_all(name, pattern = "Rev. |Dr. ", replacement = "")

```

For strings that do have a comma,  the order of the words should be reversed

```{r}
last_names_comma = str_extract_all(name1, "\\w+,")
first_names_comma = str_extract_all(name1, ", \\w.+$")

correct_order = paste(first_names_comma, last_names_comma)

```

Now insert the correctly ordered names into the main string  
```{r}
comma_positions = grep(",", name1)

name1[comma_positions] <- correct_order[comma_positions]

```

Finally, remove all commas and empty spaces  
```{r}
name1 = str_trim(str_replace_all(name1, ",", ""))
```
  
The original string and the rearranged version  
```{r}

name
name1
```

### 3 (c) Construct a logical string for second names  
  
Task: Construct a logical string indicating whether a character has a second name

Assuming one of the names (first or second) is abbreviated, use the "." as an 
indicator for a second name
```{r}
names_second_name_1 = str_detect(name1, "\\.")
```
  

An alternative approach would be to say that if a person has more than two words 
in their name, one of them should be a second name. Then we can use the count of 
space characters > 1 as and indicator.  
```{r}
names_second_name_2 = str_detect(name1, "(\\s.+){2,}")
```

In our case, the output is the same

```{r}
name1
names_second_name_1
names_second_name_2

```


## Problem 4  
  
Describe the types of strings that conform to the following regular expressions and construct an example that is matched by the regular expression.  

(a) `[0-9]+\\$`
(b) `\\b[a-z]{1,4}\\b`
(c) `.*?\\.txt$`
(d) `\\d{2}/\\d{2}/\\d{4}`
(e) `<( .+?)>.+?<``/` `\\1>` 
  
### (a) 
`[0-9]+\\$` matches a literal character `\` at the end of the string preceded by any number of digits. Example: from a string "drive123\\" this expression would match "123\\". 

The R code example does not work yet due to the issues of passing the literal character `\` into the input string:

```{r}
str_match("drive123\\", "[0-9]+\\$")

```

### (b)
`\\b[a-z]{1,4}\\b` matches any word that is lowercase and between 1 and 4 characters long 

```{r}
str_extract_all("a to few euro pound Euro","\\b[a-z]{1,4}\\b")

```

### (c)  
`.*?\\.txt$` optionally matches any number of any characters preceding a literal case-sensitive ".txt". So either any characters followed by ".txt" at the end of the string, or just ".txt". The line break character `\n` prevents backward search beyond it.
  
```{r}
str_extract_all(c("match123 ! 321.txt", "nomatch txt", "line break \n .txt"),".*?\\.txt$")

```

  
### (d)  
`\\d{2}/\\d{2}/\\d{4}` matches two digits followed by "/", then another two digits followed by "/", then four digits. This looks like a date format:  

```{r}

str_extract_all(c("10/10/1980", "10.10.1980"),"\\d{2}/\\d{2}/\\d{4}")

```
  
### (e)  
`<( .+?)>.+?<``/` `\\1>` This expression finds the shortest possible character or group of characters inside HTML tag marks `< >`, and then returns anything between this found matching group and its value enclosed in `</ >`. Thus it can extract parts of text that are inside html code:  

```{r}
str_extract_all(c("<p> This paragraph is matched </p>", "<This is not matched>"),"<(.+?)>.+?</\\1>")
```



