---
title: "Data 607 Project 2-1"
author: "by Dmitriy Vecheruk"
date: "9 October 2016"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    theme: united
    code_folding: show
    
---
### 1. Introduction

This is the first part of the Project 2.  
This exercise is devoted to analyzing the nutrional properties of different types of milk.  
The tasks are:  
  
1) Search the USDA website for five different types of milk (examples could include almond, cow, rice, goat, hemp), and combine into one master table.  
2) Manipulate the table format so that each row represents an observation for a specific type of milk. Nutritional variables should be displayed as separate fields  
3) Calculate the ratio of protein to total calories for each type of milk  
4) Calculate the average calorie content per cup across all milk types  
5) Sort the observations in ascending order by total calories per cup  
6) Calculate net carbohydrates for each milk type (i.e. gross carbohydrates - dietary fiber)  

### 2. Read the data

Unfortunately, the nutrient information for just three different types of milk (almond, cow, rice) was found on the USDA website. In particular, the analyzed cow milk has 3.7% fat, almond and rice milk drinks are unsweetened. This data is downloaded into three data frames for subsequent transformaiton.

```{r,message=FALSE,warning=FALSE}
# Load the required packages
library(readr)
library(knitr)
library(tidyr)
library(dplyr)
library(plotly)
```


```{r, cache = T, message=FALSE,warning=FALSE}

cow = read_csv("https://ndb.nal.usda.gov/ndb/foods/show/71?format=Abridged&reportfmt=csv&Qv=1", skip = 4,trim_ws = T)

almond = read_csv("https://ndb.nal.usda.gov/ndb/foods/show/4188?format=Abridged&reportfmt=csv&Qv=1", skip = 4, trim_ws = T)

rice = read_csv("https://ndb.nal.usda.gov/ndb/foods/show/4472?format=Abridged&reportfmt=csv&Qv=1", skip = 4, trim_ws = T)

```

Intially, the data is formatted so that for each product, the nutrient data is provided in rows. Below is the cow milk data as an example: 

```{r, message=FALSE,warning=FALSE}

head(cow)

```

Even though this is an example of a "long" dataset that needs to be transformed into a "wide" format, this transformation is justified, as the current format is not compliant with the tidy data norm. 
The problem with the current form is that each row contains observations of a different nutrient measured in different units, even though all of them are actually different dimensions of the same observation: nutrient composition of a given amount of milk. This means that in this input data, the notions of "wide" and "long" are used incorrectly.
Therefore we proceed with the next step.

### 3. Transform the data into tidy format
For comparison purposes, we need the data in a single table with a column of product labels, and other columns for different nutritional variables.
For better comparison, only the values per 100g of the product will be used instead of the values per cup which are not consistently defined between the products.
Further, only the following nutritional values are relevant for the analysis: energy (total calories), protein, and dietary fiber

We define a set of dplyr/tidyr transformations as a function and then apply it to each of the data sets.
```{r}

transformer = function(x, product){
  x = x %>% 
    select(Nutrient, Value = contains("100")) %>% # Filter out required rows only
    filter(complete.cases(.)) %>% # leave complete cases only
    filter(Nutrient %in% c("Energy", 
                           "Protein",
                           "Carbohydrate, by difference",
                           "Fiber, total dietary")) %>%  # Filter the required nutrients
    spread(Nutrient, Value) %>% # Transform into wide format
    mutate(Product = product) # Add product label
}
  

```

Apply the function to the datasets and bind them together
```{r, cache = T}

cow1 = transformer(cow, "cow")
rice1 = transformer(rice, "rice")
almond1 = transformer(almond, "almond")

df = bind_rows(cow1, rice1, almond1)
```


The corrected tidy format table looks as follows with the defined subset of the variables:

```{r}
library(knitr)
kable(df)

```


### 4. Calculate summary statistics

Note: instead of per cup values, per 100g values will be used for comparison.

#### 4.1. Calculate the ratio of protein to total calories for each type of milk

```{r, message=F,warning=F}

# Rename the columns for convenience
names(df) = c("Carbohydrate_g", "Energy_kkcal", "Fiber_g", "Protein_g", "Product")

# Calculate the statistic and make a plot
df = df %>% 
  mutate(pr_ratio = Protein_g/Energy_kkcal)

plot_ly(data=arrange(df, pr_ratio), y = pr_ratio, x = Product, type = "bar") %>%  layout(title = "Grams protein per kkcal of Energy", yaxis = list(title = "Protein ratio"))
```
  
We see that plant-based types of milk provide less protein than cow milk, however almond milk is much better than rice milk in this regard. 
  
#### 4.2. Calculate the average calorie content per 100g across all milk types  

The average calorie per 100g is **`r mean(df$Energy_kkcal)`** kkcal. A 250 ml cup of milk would contain on average **`r mean(df$Energy_kkcal)*2.5`** kkcal.
  
#### 4.3. Sort the observations in ascending order by total calories per 100g  

```{r}
plot_ly(data=arrange(df, Energy_kkcal), y = Energy_kkcal, x = Product, type = "bar") %>% 
  layout(title = "Energy (kkcal) per 100g", yaxis = list(title = "Total calories per 100g"))
```

In terms of the energy, cow milk is leading as well. For a low calorie option, almond milk should be preferred. 

#### 4.4. Calculate net carbohydrates for each milk type (i.e. gross carbohydrates - dietary fiber) per 100g  

```{r}
df = df %>% 
  mutate(net_carb = Carbohydrate_g - Fiber_g)

plot_ly(data=arrange(df, net_carb), y = net_carb, x = Product, 
        type = "bar", name = "Net Carb") %>% 
  add_trace(y = Carbohydrate_g, x = Product, 
            type = "bar", name = "Total Carb") %>% 
  layout(title = "Grams of total and net carbohydrates per 100g", yaxis = list(title = "Carbohydrate per 100g"))

```
  
We see that rice milk is the only type that has any dietary fiber, but it also has vastly more carbohydrates than cow and almond milk.

### 5. Conclusion
  
Overall, cow milk has more energy and more protein per 100g than plant based milk. However, among plant milk types, almond milk seems a good low-calorie, protein-rich option.
